Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-10-13 paper.tex
Add Matt's comments through p.8 (last edit incomplete)
Note, the last edit is an incomplete attempt to improve the paragraph
at the end of III.C, explaining the phenomenon of increasing <eps> at
the limb of the star. I may delete this paragraph in the end...

%\subsubsection*{Why does a covariant formulation matter?}
We formulate our ray tracing equations covariantly---free from
assumptions about the spacetime geometry or coordinates. This is essential
because we want to apply the method as a postprocessing step using
time snapshots of data computed from general relativistic evolutions.
The spacetime represented in these snapshots is not analytic (i.e. Kerr).
And even in configurations that are described well by the Kerr metric
(e.g. a low-mass disk around a massive black hole),
the evolution coordinates are unlikely to present the metric in
a familiar analytic form.
This is because integrating the Einstein Equations often requires complicated,
time-dependent gauge conditions
\citep{lind2007-gen_harmonic, fouc2013-compactness_and_spin}.

@@

We make the right hand side of Eqn.~\ref{eqn:boltzmann} explicit by writing
the source term linear in $f$:
\begin{align}
  \label{eqn:boltzmann_linear}
  \frac{d}{d\lambda}f &=
  \mathscr{E} - \mathscr{K} f \\
  \label{eqn:boltzmann_linear_s}
  &= \mathscr{K}(\mathscr{S}-f)
\end{align}
where we have introduced
$\mathscr{E}$, the invariant total emissivity, and
$\mathscr{K}$, the invariant total opacity.
These describe respectively the energy gained and the energy lost
per length traveled by the neutrino, true scalar quantities
which take identical values for all observers.
In the second form we have introduced the source function
$\mathscr{S}\equiv\mathscr{E}/\mathscr{K}$, which makes manifest the
behavior of the right hand side, driving $f$ toward $\mathscr{S}$
over a lengthscale $\mathscr{K}^{-1}$ in our affine parameter;
or from Eqn.~\ref{eqn:proper_length} below, over a proper lengthscale
$\varepsilon/\mathscr{K}$ measured by our fiducial observer.

@@

Each trajectory is parameterized by affine parameter, $\lambda$, increasing in
the direction of $\ell_\beta$. We label $\lambda=\lambda_e$ at $x^\alpha_e$,
as in Fig.~\ref{fig:affine_param}.
If we multiply Eqn.~\ref{eqn:geodesic_x} by $u_\alpha \equiv dx_\alpha/ds$,
we find the element of proper distance traversed by the
neutrino as measured by our fiducial obsever $u^\alpha$ is
\begin{equation}
  \label{eqn:proper_length}
  ds=\varepsilon \,d\lambda.
\end{equation}

@@

\subsection{The Formal Solution}
\label{ssec:rendering_eqn}
We can integrate Eqn.~\ref{eqn:boltzmann_split} directly, 
with our solution split into a boundary, absorption/emission, and
a scattering term, $f=f_{\rm bdry}+f_{\rm AE}+f_{\rm SE}$:
\begin{align}
  \label{eqn:rendering_fbdry}
  f_{\rm bdry}(\lambda,\lambda_e)
  &= f(\lambda_e) e^{-\tau(\lambda,\lambda_e)}, \\
  \label{eqn:rendering_fae}
  f_{\rm AE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm AE}(\lambda')f^{\rm eq}(\lambda'), \\
  \label{eqn:rendering_fse}
  f_{\rm SE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm SE}(\lambda')\Phi(\lambda'),
\end{align}
where the optical depth is defined,
\begin{equation}
  \label{eqn:optical_depth}
  \tau(\lambda,\lambda') \equiv \int_{\lambda'}^\lambda
  d\lambda'' \, \mathscr{K}(\lambda''),
\end{equation}
and the parameterization conventions are depicted in
Fig.~\ref{fig:affine_param}.
Note that Eqn.~\ref{eqn:optical_depth} employs the total absorption
plus scattering opacity, so that the optical depth attenuating the
integrands of Eqns.~\ref{eqn:rendering_fae} and \ref{eqn:rendering_fse}
is the total optical depth.

@@

\subsection{Numerical Implementation}
\label{ssec:numerical}
Much of our numerical implementation is borrowed from the geodesic evolution
system described in~\cite{bohn2016-code}.
We integrate Eqns.~\ref{eqn:geodesic_x} and~\ref{eqn:geodesic_p} in the form
given by \citep{hugh1994-eh_finding},
and Eqns.~\ref{eqn:rendering_fae}, \ref{eqn:rendering_fse},
and \ref{eqn:optical_depth} in the form given below.
By using the time-component of Eqn.~\ref{eqn:geodesic_x} ($dt=d\lambda \, p^t$)
we may transform our integrations to coordinate time.
The coupled system of ordinary differential equations is
\begin{align}
  \label{eqn:num_x}
  \frac{dx^i}{dt} &=
  g^{ij} \frac{p_j}{p^t} - \beta^{i},\\
  \label{eqn:num_p}
  \frac{dp_i}{dt} &=
  -\alpha \alpha_{,i} p^t
  + \beta^k_{,i} p_k
  - \frac{1}{2} g^{jk}_{,i}\frac{p_j p_k}{p^t},\\
  \label{eqn:num_tau}
  \frac{d\tau}{dt} &=
  - \frac{1}{p^t}\mathscr{K},\\
  \label{eqn:num_fae}
  \frac{df_{\rm AE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm AE}} \,f^{\rm eq},\\
  \label{eqn:num_fse}
  \frac{df_{\rm SE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm SE}} \,\Phi.
\end{align}

We integrate each ray until we reach a terminal optical depth of
$\tau_{\rm term}$ at the earliest effective emission event $x_e^\alpha$.
The concept of earliest emission event is a fictitious construct we use
to allow us to truncate our integration at an event along the ray where any
further additions to the field are negligible due to the large optical depth
between $x_e^\alpha$ and $x_o^\alpha$.
We choose $\tau_{\rm term}=14$ so that $e^{-\tau_{\rm term}}<10^{-6}$,
and we then discard the contribution of $f_{\rm bdry}$
(Eqn.~\ref{eqn:rendering_fbdry}).

@@

Fig.~\ref{fig:mfps-50km} presents the neutrino mean free paths
at this thermodynamic point over two energy decades.
For this test we set up our slab large enough for neutrinos of all
energies to be trapped.
Our domain extends to $\tau_{\rm term}\,L_{\rm mfp,max}\approx10^7\,{\rm km}$,
since our ray tracing algorithm integrates rays to terminal optical
depths of $\tau_{\rm term}\geq14$.

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t \frac{dt'}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\footnote{TODO: could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

@@

As Fig.~\ref{fig:mfps-50km} shows, integrating the distribution functions
over the energies $\varepsilon\in[0,100]$~MeV
probes our numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\footnote{TODO: discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

A stationary observer measures an energy of $\tilde{\varepsilon}$
for a neutrino with momentum
$p_\alpha \rightarrow \tilde{\varepsilon}(-1,\Omega_i)$
and direction $\Omega_i \rightarrow (\sin A, 0, \cos A)$.
In the fluid frame this neutrino has energy $\varepsilon=-u^\alpha p_\alpha$;
therefore the average energy varies with observing angle like
\begin{equation}
  \label{eqn:doppler_avg_eps}
  \langle \tilde{\varepsilon} \rangle(\cos A) =
  \langle \varepsilon \rangle^{\rm eq} \frac{1}{W(1-v \cos A)},
\end{equation}
where our symbol $\langle \tilde{\varepsilon} \rangle(\cos A)$ expresses
a functional dependence on $\cos A$.
The equilibrium average energy $\langle \varepsilon \rangle^{\rm eq}$
is given by $T \mathscr{F}_3(\eta_\nu)/\mathscr{F}_2(\eta_\nu)\approx 3T$,
with the Fermi integrals given in Eqn.~\ref{eqn:fermi_integral}.
Eqn.~\ref{eqn:doppler_avg_eps} describes the well-known Doppler effect.

@@

\subsection{Idealized Star:
  Testing the Decoupling Regime}
\label{ssec:test_ab_star}
The homogeneous configurations of the previous sections may be extended to
probe the solution outside the optically thick regime
by setting up an idealized homogeneous star, with the thermodynamic variables
constant inside radius $R$ and vanishing outside.
We choose $R=50$~km and place our observer at $r=75$~km,
at which position there is a radiation cone of half-opening angle
$\cos A_{\rm max}\approx0.75$.

The formal solutions of
Eqns.~\ref{eqn:rendering_fae} and \ref{eqn:rendering_fse}
may be directly integrated in this scenario. Assuming Minkowski spacetime
and stationary fluid we have
\begin{align}
  \label{eqn:fae_homogeneous_general}
  f_{\rm AE} &=
  \frac{\chi_a^{*}}{\chi} f^{\rm eq} \left(1-e^{-\chi s}\right) \\
  \label{eqn:fse_homogeneous_general}
  f_{\rm SE} &=
  \frac{\chi_s}{\chi} \Phi \left(1-e^{-\chi s}\right) \\
  \label{eqn:geometric_s}
  s &= 2 R \left(1-\frac{r^2}{R^2}(1-\cos^2 A)\right)^{1/2},
\end{align}
where $s$ is the path length traversed by the ray through the star,
the total opacity is $\chi=\chi_a^*+\chi_s$, and the stimulated absorption
opacity $\chi_a^*$ and elastic scattering opacity $\chi_s$
are defined in App.~\ref{sec:source_terms}.

@@

Expanding the exponential in Eqn.~\ref{eqn:fae_homogeneous_general},
we see the distribution function takes the limiting values:
\begin{equation}
  \label{eqn:fae_homogeneous_limits}
  f_{\rm AE}(\cos A) = f^{\rm eq}
  \begin{cases}
    \chi s(\cos A) & s \ll \chi^{-1}\,\textrm{optically thin}\\
    1                  & s \gg \chi^{-1}\,\textrm{optically thick}.
  \end{cases}
\end{equation}
This agrees with Fig.~\ref{fig:f_absorption_sphere}, with $\nu_x$
in the optically thin limit and $\nu_e$ in the optically thick limit.

@@

Though we do not display it here,
we also examine $\langle \tilde{\varepsilon} \rangle(\cos A)$
as in Eqn.~\ref{eqn:avg_eps_per_cosA}, and robustly reproduce the
common feature of increasing average energies at the limb of the star
for $\nu_e$ and $\bar{\nu}_e$. In contrast $\langle \tilde{\varepsilon} \rangle$
for $\nu_x$ is constant across angles, as expected when the star
is optically thin at even the highest energies $\tilde{\varepsilon}\sim100$~MeV.

>>>>

%\subsubsection*{Why does a covariant formulation matter?}
We formulate the ray tracing equations covariantly---free from
assumptions about the spacetime geometry or coordinates. This is essential
because we want to apply the method as a postprocessing step using
time snapshots of data computed from general relativistic evolutions.
The spacetime represented in these snapshots is not analytic (i.e. Kerr).
And even in configurations that are described well by the Kerr metric
(e.g. a low-mass disk around a massive black hole),
the evolution coordinates are unlikely to present the metric in
a familiar analytic form.
This is because integrating the Einstein Equations often requires complicated,
time-dependent gauge conditions
\citep{lind2007-gen_harmonic, fouc2013-compactness_and_spin}.

@@

We make the right hand side of Eqn.~\ref{eqn:boltzmann} explicit by writing
the source term linear in $f$:
\begin{align}
  \label{eqn:boltzmann_linear}
  \frac{d}{d\lambda}f &=
  \mathscr{E} - \mathscr{K} f \\
  \label{eqn:boltzmann_linear_s}
  &= \mathscr{K}(\mathscr{S}-f)
\end{align}
where we have introduced
$\mathscr{E}$, the invariant total emissivity, and
$\mathscr{K}$, the invariant total opacity.
These describe respectively the energy gained and the energy lost
per length traveled by the neutrino, true scalar quantities
which take identical values for all observers.
In the second form we have introduced the source function
$\mathscr{S}\equiv\mathscr{E}/\mathscr{K}$, which makes manifest the
behavior of the right hand side, driving $f$ toward $\mathscr{S}$
over a lengthscale $\mathscr{K}^{-1}$ in the affine parameter;
or from Eqn.~\ref{eqn:proper_length} below, over a proper lengthscale
$\varepsilon/\mathscr{K}$ measured by the fiducial observer.

@@

Each trajectory is parameterized by affine parameter, $\lambda$, increasing in
the direction of $\ell_\beta$. We label $\lambda=\lambda_e$ at $x^\alpha_e$,
as in Fig.~\ref{fig:affine_param}.
If we multiply Eqn.~\ref{eqn:geodesic_x} by $u_\alpha \equiv dx_\alpha/ds$,
we find the element of proper distance traversed by the
neutrino as measured by the fiducial observer $u^\alpha$ is
\begin{equation}
  \label{eqn:proper_length}
  ds=\varepsilon \,d\lambda.
\end{equation}

@@

\subsection{The Formal Solution}
\label{ssec:rendering_eqn}
We can integrate Eqn.~\ref{eqn:boltzmann_split} directly, 
with the solution split into a boundary, absorption/emission, and
a scattering term, $f=f_{\rm bdry}+f_{\rm AE}+f_{\rm SE}$:
\begin{align}
  \label{eqn:rendering_fbdry}
  f_{\rm bdry}(\lambda,\lambda_e)
  &= f(\lambda_e) e^{-\tau(\lambda,\lambda_e)}, \\
  \label{eqn:rendering_fae}
  f_{\rm AE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm AE}(\lambda')f^{\rm eq}(\lambda'), \\
  \label{eqn:rendering_fse}
  f_{\rm SE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm SE}(\lambda')\Phi(\lambda'),
\end{align}
where the optical depth is defined,
\begin{equation}
  \label{eqn:optical_depth}
  \tau(\lambda,\lambda') \equiv \int_{\lambda'}^\lambda
  d\lambda'' \, \mathscr{K}(\lambda''),
\end{equation}
and the parameterization conventions are depicted in
Fig.~\ref{fig:affine_param}.
Note that Eqn.~\ref{eqn:optical_depth} employs the total absorption
plus scattering opacity, so that the optical depth attenuating the
integrands of Eqns.~\ref{eqn:rendering_fae} and \ref{eqn:rendering_fse}
is the total optical depth.

@@

\subsection{Numerical Implementation}
\label{ssec:numerical}
Much of our numerical implementation is borrowed from the geodesic evolution
system described in~\cite{bohn2016-code}.
We integrate Eqns.~\ref{eqn:geodesic_x} and~\ref{eqn:geodesic_p} in the form
given by \citep{hugh1994-eh_finding},
and Eqns.~\ref{eqn:rendering_fae}, \ref{eqn:rendering_fse},
and \ref{eqn:optical_depth} in the form given below.
By using the time-component of Eqn.~\ref{eqn:geodesic_x} ($dt=d\lambda \, p^t$)
we may transform the integrations to coordinate time.
The coupled system of ordinary differential equations is
\begin{align}
  \label{eqn:num_x}
  \frac{dx^i}{dt} &=
  g^{ij} \frac{p_j}{p^t} - \beta^{i},\\
  \label{eqn:num_p}
  \frac{dp_i}{dt} &=
  -\alpha \alpha_{,i} p^t
  + \beta^k_{,i} p_k
  - \frac{1}{2} g^{jk}_{,i}\frac{p_j p_k}{p^t},\\
  \label{eqn:num_tau}
  \frac{d\tau}{dt} &=
  - \frac{1}{p^t}\mathscr{K},\\
  \label{eqn:num_fae}
  \frac{df_{\rm AE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm AE}} \,f^{\rm eq},\\
  \label{eqn:num_fse}
  \frac{df_{\rm SE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm SE}} \,\Phi.
\end{align}

We integrate each ray until we reach a terminal optical depth of
$\tau_{\rm term}$ at the earliest effective emission event $x_e^\alpha$.
The concept of earliest emission event is a fictitious construct we use
to allow us to truncate the integration at an event along the ray where any
further additions to the field are negligible due to the large optical depth
between $x_e^\alpha$ and $x_o^\alpha$.
We choose $\tau_{\rm term}=14$ so that $e^{-\tau_{\rm term}}<10^{-6}$,
and we then discard the contribution of $f_{\rm bdry}$
(Eqn.~\ref{eqn:rendering_fbdry}).

@@

Fig.~\ref{fig:mfps-50km} presents the neutrino mean free paths
at this thermodynamic point over two energy decades.
For this test we choose a slab large enough for neutrinos of all
energies to be trapped.
We sample the distribution function with ray tracing over a uniform grid
of 40 points in energy $\varepsilon \in [0,100]$~MeV.
The domain extends to $\tau_{\rm term}\,L_{\rm mfp,max}\approx10^7\,{\rm km}$,
since our ray tracing algorithm integrates rays to terminal optical
depths of $\tau_{\rm term}\geq14$.

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}, at the single energy
$\varepsilon=11.25$~MeV. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t \frac{dt'}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\footnote{TODO: could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

@@

As Fig.~\ref{fig:mfps-50km} shows, integrating the distribution functions
over the energies $\varepsilon\in[0,100]$~MeV
probes the numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\footnote{TODO: discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

A stationary observer measures an energy of $\tilde{\varepsilon}$
for a neutrino with momentum
$p_\alpha \rightarrow \tilde{\varepsilon}(-1,\Omega_i)$
and direction $\Omega_i \rightarrow (\sin A, 0, \cos A)$.
In the fluid frame this neutrino has energy $\varepsilon=-u^\alpha p_\alpha$;
therefore the average energy varies with observing angle like
\begin{equation}
  \label{eqn:doppler_avg_eps}
  \langle \tilde{\varepsilon} \rangle(\cos A) =
  \langle \varepsilon \rangle^{\rm eq} \frac{1}{W(1-v \cos A)},
\end{equation}
where the symbol $\langle \tilde{\varepsilon} \rangle(\cos A)$ expresses
a functional dependence on $\cos A$.
The equilibrium average energy $\langle \varepsilon \rangle^{\rm eq}$
is given by $T \mathscr{F}_3(\eta_\nu)/\mathscr{F}_2(\eta_\nu)\approx 3T$,
with the Fermi integrals given in Eqn.~\ref{eqn:fermi_integral}.
Eqn.~\ref{eqn:doppler_avg_eps} describes the well-known Doppler effect.

@@

\subsection{Idealized Star:
  Testing the Decoupling Regime}
\label{ssec:test_ab_star}
The homogeneous configurations of the previous sections may be extended to
probe the solution outside the optically thick regime
by setting up an idealized homogeneous star, with the thermodynamic variables
constant inside radius $R$ and vanishing outside.
We choose $R=50$~km and place the observer at $r=75$~km,
at which position there is a radiation cone of half-opening angle
$\cos A_{\rm max}\approx0.75$.

The formal solutions of
Eqns.~\ref{eqn:rendering_fae} and \ref{eqn:rendering_fse}
may be directly integrated in this scenario. Assuming Minkowski spacetime
and stationary fluid we have
\begin{align}
  \label{eqn:fae_homogeneous_general}
  f_{\rm AE} &=
  \frac{\chi_a^{*}}{\chi} f^{\rm eq} \left(1-e^{-\chi s}\right), \\
  \label{eqn:fse_homogeneous_general}
  f_{\rm SE} &=
  \frac{\chi_s}{\chi} \Phi \left(1-e^{-\chi s}\right),
\end{align}
where $s$ is the path length traversed by the ray through the star,
\begin{equation}
  \label{eqn:geometric_s}
  s = 2 R \left(1-\frac{r^2}{R^2}(1-\cos^2 A)\right)^{1/2}.
\end{equation}
The total opacity is $\chi=\chi_a^*+\chi_s$, and the stimulated absorption
opacity $\chi_a^*$ and elastic scattering opacity $\chi_s$
are defined in App.~\ref{sec:source_terms}.

@@

We can explain these features quantitatively by examining the limits
of Eqn.~\ref{eqn:fae_homogeneous_general}, expanding the exponential function
in powers of $\chi s$; the distribution function takes the limiting values
\begin{equation}
  \label{eqn:fae_homogeneous_limits}
  f_{\rm AE}(\cos A) = f^{\rm eq}
  \begin{cases}
    \chi s(\cos A) & s \ll \chi^{-1}\,\textrm{(thin)}\\
    1                  & s \gg \chi^{-1}\,\textrm{(thick)}.
  \end{cases}
\end{equation}
These limits are represented in Fig.~\ref{fig:f_absorption_sphere}:
with $\nu_x$ in the optically thin limit at all viewing angles,
and $\nu_e$ in the optically thick limit at viewing angles
$\cos A \gtrsim 0.8$.

@@

Though we do not display it here,
we also examine $\langle \tilde{\varepsilon}\rangle$ as a function of
viewing angle (via Eqn.~\ref{eqn:avg_eps_per_cosA}).
In the interior of the star
$\langle \tilde{\varepsilon_{\nu_x}}\rangle$ is constant across viewing angle,
\footnote{TODO: need to finish cleaning up this description}
and find the counterintuitive
feature of increasing average energies at the limb of the star
for $\nu_e$ and $\bar{\nu}_e$.
In contrast $\langle \tilde{\varepsilon} \rangle$
for $\nu_x$ is constant across angles, as expected when the star
is optically thin at even the highest energies $\tilde{\varepsilon}\sim100$~MeV.

For viewing angles $\cos A \sim 0.75$, at the limb of the star,
rays pass through significantly less material, i.e.\ $s \ll R$.
Because of the opacity's approximate $\varepsilon^2$ dependence,
the highest-energy rays saturate to $f^{\rm eq}$, increasing the average spectral energy.
